Generating Images From Captions\\ With Attention

### 2015-10-26 iclr-paper-new.tex
More changes made

Unlike the original DRAW network where latent variables are independent unit Gaussians $\mathcal{N}(0, I)$, the latent variables in the proposed alignDRAW model have their mean and variance depend on the previous recurrent hidden states $\hdec_{t-1}$ as in \citep{bachman_sdm}, except for $\lat_1 \sim \prior(\Lat_1) = \mathcal{N}(0, I)$. It has been shown in \citep{bachman_sdm} that model performance is improved by including dependencies between latent variables. Formally, the image is generated by iteratively computing the following equations for $t=1,...,T$ (see Figure~\ref{fig:figmodel}):
\begin{align}
\label{eq:x_hat}
\lat_t &\sim \prior(\Lat_t|\Lat_{1:t-1}) = \mathcal{N}(\mu_{t}(\hdec_{t-1}), \sigma_{t}(\hdec_{t-1})),\\
\hdec_t &= \decoder(\hdec_{t-1}, z_t, s_{t-1}), \\
s_{t} &= align(\hdec_{t-1}, \hlangall), \\
\canv_t &= \canv_{t-1} + \writeop(\hdec_t), 
\label{eq:write}
%\canv_t &= \canv_{t-1} + \writeop(\hdec_t)
\end{align}
where $\writeop$ is the same attention operator as in \citep{gregor_draw}.
The $align$ function is used to compute the alignment between the input caption and intermediate image generative steps \citep{bahdanau_mt}. Given the caption representation from the language model, $\hlangall = [\hlang_{1}, \hlang_{2}, ..., \hlang_{N}]$, the $align$ operator outputs a dynamic sentence representation $s_t$ at each step through a weighted sum using alignment probabilities $\alpha_{1...N}$:
%\begin{align}
%\vspace{-0.1in}
$s_t=align(\hdec_{t-1}, \hlangall) = \alpha_{1}\hlang_{1} + \alpha_{2}\hlang_{2} + ... + \alpha_{N}\hlang_{N}.$
%\end{align}
The corresponding alignment probabilities $\alpha_{1...N}$ at each step are obtained using the caption representation $\hlangall$ and the hidden state of the generative model $\hdec_{t}$, as in \citep{bahdanau_mt}. The final canvas $\canv_T$ is transformed using a sigmoid function to produce an Bernoulli data likelihood over pixels given the latent variables and the input caption.

@@

%The model is learned by the modified version of Stochastic Gradient Variation Bayes (SGVB) algorithm introduced by \cite{kingma_vae}. 
The model is trained to maximize a variational lower bound $\loss$ on the marginal likelihood of the correct image $\oimage$ given the input caption $\icaption$, where $\loss = \sum_{\Lat}Q(\Lat\given\oimage,\icaption)\left(\log P(\oimage\given\icaption, \Lat_{1:T}) + \kldiv\left(Q(\Lat\given\oimage,\icaption)\klBars P(\Lat)\right)\right) \le \log P(\oimage\given\icaption)$ The posterior inference over the latent variables $\Lat_{1:T}$ is approximated using a inference recurrent neural net. Similarly to the DRAW network, the inference RNN produces an approximate posterior $Q(\Lat_{1:T}\given\oimage,\icaption)$ via a $\readop$ operator over the training image at each step. 
\begin{align}
\label{eq:x_hat}
\hat{\oimage}_t &= \oimage-\sigmoid(\canv_{t-1})\\
\label{eq:read}
r_t &= \readop(\oimage_t, \hat{\oimage}_t, \hdec_{t-1})\\
\henc_t &= \encoder(\henc_{t-1}, [r_t, \hdec_{t-1}])\\
\post(\Lat_t|\oimage,\icaption,\Lat_{1:t-1}) &= \mathcal{N}\left(\mu_t(\henc_t), \sigma_t(\henc_t)\right)
\end{align}

@@

\section{Experiments}
\subsection{MNIST With Captions}
As a first experiment, we trained our proposed model on the MNIST dataset with artificial captions. Either one or two digits from the MNIST training dataset were placed on a $60 \times 60$ blank image. One digit was placed in one of the four (top-left, top-right, bottom-left or bottom-right) corners of the image. Two digits were either placed horizontally or vertically in nonoverlapping fashion. 
While many generative models were trained on the binarized version of the MNIST dataset, we trained directly on pixel intensities with a binary cross-entropy cost function. 
The generated images with the attention alignments are displayed in Figure~\ref{fig:figmnist}.

@@

\begin{figure}[!b]
\captionsetup[subfigure]{labelformat=empty}
\vspace{-0.5cm}
\begin{center}
\subfloat[]{\includegraphics[width=0.23\textwidth]{figures/new/a-very-large-commercial-plane-flying-in-blue-skies-closest.png}}\quad
%
\subfloat[]{\includegraphics[width=0.23\textwidth]{figures/new/a-very-large-commercial-plane-flying-in-rainy-skies-closest.png}}\quad
%
\subfloat[]{\includegraphics[width=0.23\textwidth]{figures/new/a-herd-of-elephants-walking-across-a-dry-grass-field-closest.png}}\quad
%
\subfloat[]{\includegraphics[width=0.23\textwidth]{figures/new/a-herd-of-elephants-walking-across-a-green-grass-field-closest.png}}\\
\vspace{-0.45cm}
%
\subfloat[A very large commercial plane flying in \underline{blue} skies.
]{\includegraphics[width=0.23\textwidth]{figures/new/a-very-large-commercial-plane-flying-in-blue-skies-sharp.png}}\quad
%
\subfloat[A very large commercial plane flying in \underline{rainy} skies.
]{\includegraphics[width=0.23\textwidth]{figures/new/a-very-large-commercial-plane-flying-in-rainy-skies-sharp.png}}\quad
%
\subfloat[A herd of elephants walking across a \underline{dry} grass field.
]{\includegraphics[width=0.23\textwidth]{figures/new/a-herd-of-elephants-walking-across-a-dry-grass-field-sharp.png}}\quad
%
\subfloat[A herd of elephants walking across a \underline{green} grass field.
]{\includegraphics[width=0.23\textwidth]{figures/new/a-herd-of-elephants-walking-across-a-green-grass-field-sharp.png}}\quad
%
\end{center}
\caption{\textbf{Bottom}: Examples of changing the background while keeping the caption fixed. \textbf{Top}: The respective nearest training images based on pixel-wise L2 distance. The nearest images from the training set also indicate that the model was not simply copying the patterns it observed during the learning phase.}
\label{fig:genimages2}
\vspace{-0.3cm}
\end{figure}

@@

Instead of calculating error per pixel, we turn to a smarter metric, the Structural Similarity Index (SSI) \citep{wang_ssi}, which incorporates luminance and contrast masking into the error calculation. Strong inter-dependencies of closer pixels is also taken into account and the metric is calculated on small windows of the images. Due to independence property of test captions, we sampled fifty images from the prior of each generative model for every caption in the test set in order to calculate SSI. The results are reported in Table~\ref{tab:results}.

@@

We showed that our proposed model managed to generate reasonable images corresponding to respective captions as well as managed to generalize to completely novel scenarios which are very unlikely to occur in real life. Sequential generation of images using attention played a crucial part in it. However the conditional DRAW model still underfits on training data which leads to the blurry samples. We think that extending this process by generating very small sections of images over potentially hundreds of timesteps might help the model fit the data and generate sharper images.

@@

\bibliography{iclr-paper}
\bibliographystyle{iclr-style/iclr2016_conference}

>>>>

Unlike the original DRAW network where latent variables are independent unit Gaussians $\mathcal{N}(0, I)$, the latent variables in the proposed alignDRAW model have their mean and variance depend on the previous recurrent hidden states $\hdec_{t-1}$ as in \citep{bachman_sdm}, except for $\lat_1 \sim \prior(\Lat_1) = \mathcal{N}(0, I)$. It has been shown in \citep{bachman_sdm} that model performance is improved by including dependencies between latent variables. Formally, the image is generated by iteratively computing the following equations for $t=1,...,T$ (see Figure~\ref{fig:figmodel}):
\begin{align}
\label{eq:x_hat}
\lat_t &\sim \prior(\Lat_t|\Lat_{1:t-1}) = \mathcal{N}(\mu_{t}(\hdec_{t-1}), \sigma_{t}(\hdec_{t-1})),\\
\hdec_t &= \decoder(\hdec_{t-1}, [z_t, s_{t-1}]), \\
s_{t} &= align(\hdec_{t-1}, \hlangall), \\
\canv_t &= \canv_{t-1} + \writeop(\hdec_t), 
\label{eq:write}
%\canv_t &= \canv_{t-1} + \writeop(\hdec_t)
\end{align}
where $\writeop$ is the same attention operator as in \citep{gregor_draw}. The $\writeop$ operator applies an array of 2D Gaussian filters on the patch that places it on the canvas with smoothly varying location and zoom.
The $align$ function is used to compute the alignment between the input caption and intermediate image generative steps \citep{bahdanau_mt}. Given the caption representation from the language model, $\hlangall = [\hlang_{1}, \hlang_{2}, ..., \hlang_{N}]$, the $align$ operator outputs a dynamic sentence representation $s_t$ at each step through a weighted sum using alignment probabilities $\alpha_{1...N}$:
%\begin{align}
%\vspace{-0.1in}
$s_t=align(\hdec_{t-1}, \hlangall) = \alpha_{1}\hlang_{1} + \alpha_{2}\hlang_{2} + ... + \alpha_{N}\hlang_{N}.$
%\end{align}
The corresponding alignment probabilities $\alpha_{1...N}$ at each step are obtained using the caption representation $\hlangall$ and the hidden state of the generative model $\hdec_{t}$, by first calculating energies $e_{1...N}$ using $tanh$ perceptron and then normalizing them, as in \citep{bahdanau_mt}:
%\begin{align}
$\alpha_{k}=\frac{exp(v^{T}tanh(U\hlang_{k} + W\hdec_{t} + b)}{\sum_{j=1}^{N}exp(v^{T}tanh(U\hlang_{j} + W\hdec_{t} + b)}.$
%\end{align}
The final canvas $\canv_T$ is transformed using a sigmoid function to produce an Bernoulli data likelihood over pixels given the latent variables and the input caption.

@@

%The model is learned by the modified version of Stochastic Gradient Variation Bayes (SGVB) algorithm introduced by \cite{kingma_vae}. 
The model is trained to maximize a variational lower bound $\loss$ on the marginal likelihood of the correct image $\oimage$ given the input caption $\icaption$, where $\loss = \sum_{\Lat}Q(\Lat\given\oimage,\icaption)\left(\log P(\oimage\given\icaption, \Lat_{1:T}) + \kldiv\left(Q(\Lat\given\oimage,\icaption)\klBars P(\Lat)\right)\right) \le \log P(\oimage\given\icaption)$. The posterior inference over the latent variables $\Lat_{1:T}$ is approximated using a inference recurrent neural net. Similarly to the DRAW network, the inference RNN produces an approximate posterior $Q(\Lat_{1:T}\given\oimage,\icaption)$ via a $\readop$ operator, which reads a patch from image using an array of Gaussian filters (inverse of $\writeop$), over the training image at each step. 
\begin{align}
\label{eq:x_hat}
\hat{\oimage}_t &= \oimage-\sigmoid(\canv_{t-1})\\
\label{eq:read}
r_t &= \readop(\oimage_t, \hat{\oimage}_t, \hdec_{t-1})\\
\henc_t &= \encoder(\henc_{t-1}, [r_t, \hdec_{t-1}])\\
\post(\Lat_t|\oimage,\icaption,\Lat_{1:t-1}) &= \mathcal{N}\left(\mu_t(\henc_t), \sigma_t(\henc_t)\right)
\end{align}

@@

\newpage
\section{Experiments}
\subsection{MNIST With Captions}
As a first experiment, we trained our proposed model on the MNIST dataset with artificial captions. Either one or two digits from the MNIST training dataset were placed on a $60 \times 60$ blank image. One digit was placed in one of the four (top-left, top-right, bottom-left or bottom-right) corners of the image. Two digits were either placed horizontally or vertically in nonoverlapping fashion. 
While many generative models were trained on the binarized version of the MNIST dataset, we trained directly on pixel intensities with a binary cross-entropy cost function.

The generated images together with the attention alignments are displayed in Figure~\ref{fig:figmnist}. The model correctly displayed defined digits at the described positions and even managed to generalize reasonably to the configurations that were never present during training. In case of generating two digits, the model attended to the digit in the caption that was drawn at the respective timestep while completely ignoring the other digit. As for generating one digit, the model attended to the digit in the caption during the whole generation process. However in both cases, the model placed small attention values on the words describing the position of digits in the images.

@@

\begin{figure}[!t]
\captionsetup[subfigure]{labelformat=empty}
%\vspace{-0.5cm}
\begin{center}
\subfloat[]{\includegraphics[width=0.23\textwidth]{figures/new/a-very-large-commercial-plane-flying-in-blue-skies-closest.png}}\quad
%
\subfloat[]{\includegraphics[width=0.23\textwidth]{figures/new/a-very-large-commercial-plane-flying-in-rainy-skies-closest.png}}\quad
%
\subfloat[]{\includegraphics[width=0.23\textwidth]{figures/new/a-herd-of-elephants-walking-across-a-dry-grass-field-closest.png}}\quad
%
\subfloat[]{\includegraphics[width=0.23\textwidth]{figures/new/a-herd-of-elephants-walking-across-a-green-grass-field-closest.png}}\\
\vspace{-0.45cm}
%
\subfloat[A very large commercial plane flying in \underline{blue} skies.
]{\includegraphics[width=0.23\textwidth]{figures/new/a-very-large-commercial-plane-flying-in-blue-skies-sharp.png}}\quad
%
\subfloat[A very large commercial plane flying in \underline{rainy} skies.
]{\includegraphics[width=0.23\textwidth]{figures/new/a-very-large-commercial-plane-flying-in-rainy-skies-sharp.png}}\quad
%
\subfloat[A herd of elephants walking across a \underline{dry} grass field.
]{\includegraphics[width=0.23\textwidth]{figures/new/a-herd-of-elephants-walking-across-a-dry-grass-field-sharp.png}}\quad
%
\subfloat[A herd of elephants walking across a \underline{green} grass field.
]{\includegraphics[width=0.23\textwidth]{figures/new/a-herd-of-elephants-walking-across-a-green-grass-field-sharp.png}}\quad
%
\end{center}
\caption{\textbf{Bottom}: Examples of changing the background while keeping the caption fixed. \textbf{Top}: The respective nearest training images based on pixel-wise L2 distance. The nearest images from the training set also indicate that the model was not simply copying the patterns it observed during the learning phase.}
\label{fig:genimages2}
\vspace{-0.3cm}
\end{figure}

@@

Instead of calculating error per pixel, we turn to a smarter metric, the Structural Similarity Index (SSI) \citep{wang_ssi}, which incorporates luminance and contrast masking into the error calculation. Strong inter-dependencies of closer pixels is also taken into account and the metric is calculated on small windows of the images. Due to independence property of test captions, we sampled fifty images from the prior of each generative model for every caption in the test set in order to calculate SSI. The results are reported on Table~\ref{tab:results}.

@@

We showed that our proposed model managed to generate reasonably looking images corresponding to respective captions as well as managed to generalize to completely novel scenarios which are very unlikely to occur in real life. Sequential generation of images over some number of steps has played an important part in this process. However, the tendency of the conditional DRAW models to underfit on the training data has resulted in the blurry samples. As a result, the samples were sharpened by the GAN trained on the residuals of the images. Unfortunately, this is not a quite satisfying solution due to the fact that the whole model wasn't trained in the end-to-end fashion. We think that further extention of the capabilities of the sequential generation of images might help the variational models better fit the data and escape the curse of blurriness.

@@

\newpage
\bibliography{iclr-paper}
\bibliographystyle{iclr-style/iclr2016_conference}

### 2015-10-26 supp.tex
fixed small error in supplement
More changes made

Each parameter in alignDRAW was initialized by sampling from the Gaussian distribution of mean $0$ and standard deviation of $0.01$. All models were trained with \textit{RMSProp} with initial learning rate of $0.001$. The norm of the gradients was rescaled to $10.0$ during training. For all tasks we used $128$ units in language model and $512$ units in alignment multi-layer perceptron.

% \begin{table}[!h]
% \begin{center}
% \begin{tabulary}{\linewidth}{c | c c c}
% \hline
% \multicolumn{4}{c}{\textbf{Language Model}} \\
% \hline
% Task & Vocab size & LSTM \#$h$ & Align \#$h$\\
% \hline
% MNIST With Captions & 22 & 128 & 512  \\
% Microsoft COCO & 25323 & 128 & 512
% \end{tabulary}
% \end{center}
% \end{table}

\begin{table}[!h]
\begin{center}
\begin{tabulary}{\linewidth}{c | c c c c c c}
\hline
\multicolumn{7}{c}{\textbf{Image Model}} \\
\hline
Task & \#glimpses & Inference \#$h$ & Generative \#$h$ & \#$z$ & Read Size & Write Size\\
\hline
MNIST Captions & 32 & 550 & 550 & 275 & 9 & 9\\
MS COCO & 32 & 300 & 300 & 150 & 8 & 8\\
\end{tabulary}
\end{center}
\end{table}

\subsection{Evaluation}
\begin{table}[!h]
\begin{center}
\begin{tabulary}{\linewidth}{c | c c c c}
\hline
Model & Train NLL & Valid NLL & Test NLL & Test NLL (after sharpening)\\
\hline
skipthoughtDRAW & 1794.29 & 1797.41 & 1791.37 & 2045.84 \\
noalignDRAW & 1792.14 & 1796.94 & 1791.15 & 2051.07 \\
alignDRAW & 1792.15 & 1797.24 & 1791.53 & 2042.31
\end{tabulary}
\end{center}
\end{table}

>>>>

Each parameter in alignDRAW was initialized by sampling from the Gaussian distribution of mean $0$ and standard deviation of $0.01$. The model was trained using \textit{RMSProp} with initial learning rate of $0.001$. The norm of the gradients was clipped to $10.0$ during training to deal with exploding gradients. For all tasks we used $128$ units in the language model and $512$ units in the $tanh$ perceptron in the alignment function. For the Microsoft COCO task, the alignDRAW model was trained for $18$ epochs (the model converges after $14$ epochs) and the learning rate was reduced by the factor of $0.1$ after $11$ epochs. The architectural configurations of alignDRAW models are shown on Table~\ref{tab:drawhyper}.

\begin{table}[!h]
\begin{center}
\begin{tabulary}{\linewidth}{c | c c c c c c}
\hline
\multicolumn{7}{c}{\textbf{alignDRAW Model}} \\
\hline
Task & \#glimpses & Inference \#$h$ & Generative \#$h$ & \#$z$ & Read Size & Write Size\\
\hline
Microsoft COCO & 32 & 550 & 550 & 275 & 9 & 9\\
MNIST Captions & 32 & 300 & 300 & 150 & 8 & 8\\
\end{tabulary}
\end{center}
\label{tab:drawhyper}
\caption{The architectural configurations of alignDRAW models.}
%\vspace{-0.3cm}
\end{table}

The GAN model trained on residuals which was used for sharpening had the same configuration as the $28 \times 28$ model trained by \cite{denton_lapgan} on residuals of CIFAR dataset, that can be found at \url{https://gist.github.com/soumith/e3f722173ea16c1ea0d9}. The model was trained for $6$ epochs.

\subsection{Evaluation}

On Table~\ref{tab:nll}, you can see that the alignDRAW model didn't overfit on the training data as well as managed to achieve a better lower bound of log likelihood on the test data. Also notice that, the results worsened dramatically after sharpening test images.

\begin{table}[!h]
\begin{center}
\begin{tabulary}{\linewidth}{c | c c c c}
\hline
Model & Train LL & Valid LL & Test LL & Test LL (after sharpening)\\
\hline
skipthoughtDRAW & -1794.29 & -1797.41 & -1791.37 & -2045.84 \\
noalignDRAW & -1792.14 & -1796.94 & -1791.15 & -2051.07 \\
alignDRAW & -1792.15 & -1797.24 & -1791.53 & -2042.31
\end{tabulary}
\end{center}
\label{tab:nll}
\caption{The lower bound of log likelihood of conditional DRAW models on the Microsoft COCO.}
\end{table}
